Assessment of forestry road network projection methods using the “roads” package in R

### 2022-07-19 analysis/paper/paper.Rmd
working on paper text and figures. Added new map

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

@@

# Introduction[a]

@@

Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada’s forested area. In British Columbia, over 600 000 kilometers of resource roads make up the vast majority of the province’s total road network (83%)  (Environmental Reporting BC, 2018; Forest Practices Board, 2015). Over 75% of resource roads in the province were built by the forest industry (Forest Practices Board, 2015). Resource extraction activities provide jobs and renewable raw materials (Canada, 2013b), but they also can create threats for wildlife. For example, the main threats to woodland caribou, a species at risk in Canada, are human disturbances such as forest harvesting, oil and gas exploration/extraction, and road networks. These particular disturbances fragment their habitat, and create open areas and extensive young forests that attract other species such as moose and deer which in turn, attracts an increased number of predators, such as wolves and even poachers (Canada, 2013a). Woodland caribou are often considered to be an umbrella species, which means that protecting and restoring their habitat would benefit other species also dependent on old forests, such as fishers, pine martens, and many species of birds (Mulvihill, 2018).

@@

Forest simulation models generally either fall into two categories: deterministic and stochastic (Boychuk et al., 2009; Helmes & Stockbridge, 2011). Stochastic simulation approaches use probability distributions in combination with random number generators to determine events (He & Mladenoff, 1999). Despite the importance of roads on the environment, previous spatial stochastic harvest simulation models rarely include forest road network development in their projections. The “roads” package in R aims to fill this gap in landscape modelling by projecting large-scale forest road networks based on future harvest cut blocks, an existing road network, and a road cost surface. This paper uses the “roads” package to run forestry road network simulations in British Columbia from approximately 1990 to present and compares the results to the observed modern-day forest road networks in order to assess the performance of the package and to showcase its usefulness within forest simulations.

(Introduction round 2: road context)
Several papers have previously introduced forest road network projections (Anderson & Nelson, 2004; Clark et al., 2000; Dean, 1997) however many of these methods are complex with many input parameters and are not readily available for public use. These previous projections have mainly been developed in order to facilitate operational forest road network planning and therefore require precise data and many input parameters in order to obtain the most effective network. These projection methods are therefore not very applicable for projecting forest roads over large-scale landscapes. On the other hand, many spatial stochastic forest management simulations haven’t included forest road development at all (citations). One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.). There is therefore a need in large scale landscape simulations for a tool that can adequately project forest road development with less complexity and minimal input parameters that can easily be incorporated into other complex simulations in large scale landscapes. As many ecological forecasts use increasingly complex methods and data, forecasting tools (e.g., models) should also be scalable and leverage existing models (Bodner et al., in press) (“Bridging the divide between ecological forecasts and environmental decision-making”).

@@

# Methods
## Study Area
We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as our study area because forestry is the dominant resource extraction activity in this region and much of the TSA’s total road network is composed of forestry resource roads built to gain access to forest harvest cutblocks. This allows us to project road building based on forest harvest data. With approximately 55% of Revelstoke TSA’s area being non-forested land (alpine, lakes, swamp, brush, rock, etc) (Brown & Hachey, 2010), the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development. The Revelstoke TSA is in the interior-wet belt of the province and its forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%). The TSA contains wildlife species that are at risk due to their declining populations and are believed to be negatively affected by road development, such as members of the southern mountain woodland caribou population (Brown & Hachey, 2010). The Revelstoke TSA is an example of a landscape where road projections could be used to manage critical wildlife habitats. The British Columbia TSA boundary shapefile layer was downloaded from the British Columbia Data Catalogue (Government of British Columbia, 2021a).

## Road network projection methods
Various optimization and simulation techniques have been used for projecting forestry roads into the future and across landscapes, however, we focused on the simulation methods that link multiple harvest units (a.k.a cutblocks) to an existing road network for strategic purposes. Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network. We used three methods for solving the MTAP in the roads package. These methods represent a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by the least cost paths approach (LCP) and finally, the minimum spanning tree approach (MST). In the remainder of this paper, we will focus on the LCP and MST approaches given their assumptions are well aligned with how forestry professionals would strategically design road networks.

LCP approach solves the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP). For each STAP, a least cost path from a single source to sink was solved using Dijkstra’s shortest path algorithm (1959). Dijkstra’s algorithm minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel (the details of this process are described later in the section “weighted graph”). Each STAP was then solved sequentially, with the newly projected roadway updating the existing road network and subsequently the costs of road construction. This method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP, however the projected network is sensitive to the ordering of the STAP. They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.

The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities. Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area. Our MST approach has three steps towards evaluating branch locations while projecting roads: 1) estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network, 2) creating a weighted graph where each node is either a landing or point on the existing road network and each edge or arc is weighted by the costs estimated in the previous step and 3) solving the minimum spanning tree for this graph. The minimum spanning tree was solved using Kruskal’s algorithm (1956) where every node in the graph was linked to form a tree such that the sum of edge weights is minimized, and the links do not form a loop. This method was most similar to Clark et al. (2000) approach which used a minimum spanning tree to project simple road networks during harvest scheduling. An important difference was that they based the edge weights on spatial Euclidean distance, whereas the MST approach used the costs estimated from least cost paths.

## Weighted graph
A mathematical weighted graph consisting of nodes and edges was used to parameterize either Kruskal’s or Dijkstra’s algorithms. When solving least cost paths using Dijkstra’s algorithm, each node in the graph corresponded to a pixel in a raster of our study area (100m x 100m resolution). Whereas, in the graph used for solving the minimum spanning tree via Kruskal’s algorithm, each node represented either a landing or a point on the existing road network. Therefore, a node can be represented by any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point. The landing nodes represent log decks or sources of timber and the point on the existing road is the sink where the timber will end up. The intermediate points are plausible locations to construct a road and are thus free of barriers like major water bodies. The edges connecting the nodes follow a “queens” case or allow a maximum of eight possible edges for a given node. Thus, our methods utilize a grid pattern which does not support curvature or switchbacks in road design as required in operational road design, thus we ascertain that our road projection methods are for strategic modelling purposes only. Edges are assigned weights which penalize the algorithm to account for various impedance and road construction costs. When solving least cost paths, a cost surface raster was used. The pixels of the cost surface raster corresponded to nodes in the weighted graph and thus the edge weight was the average cost between the two nodes or pixels. When solving the minimum spanning tree, the edge weight corresponded to the total cost of the least cost path between the two nodes.

## Data
To spatially define the cost of constructing and traveling on a forestry road we developed a cost surface raster. We used several spatial raster layers describing the roading cost structure, barriers to road development and road impedance. For intermediate nodes, the roading cost structure followed the Interior Appraisal Manual of BC (source) where the cost per kilometer of road was a linear function of slope. Separate costs for river and pipeline crossings were then added to the roading cost. Barriers for road development masked the cost surface raster and these included the major water bodies of BC. Lastly, to coerce the algorithm to follow existing roads, all nodes representing the existing road network were assigned the smallest cost (Table 1). The cost surface for Revelstoke TSA consisted of 1703936 pixels with a resolution of 100 meters by 100 meters.

Table 1. Definitions of road building costs for building the cost surface. 
```{r}
tbl <- "Cost Type|Road Attribute|Value ($ per km)|Source
Barriers of road development|Major lakes|infinity|Major water bodies of BC
Impedance|Existing roads|0.01|Assumed
Road Construction Cost Structure|Pipe crossing|1911|Interior Appraisal Manual
|River crossing|2130|Interior Appraisal Manual
|Base cost|16178 + 504*slope|Interior Appraisal Manual
"
df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df) <- unname(as.list(df[1,]))
df <- df[-1,] # remove first row
row.names(df)<-NULL
knitr::kable(df)
```

The road network used was a combination of the digital road atlas (DRA) (Data BC, 2021) and forest tenure road (FLNRORD, 2021b) datasets from the Government of British Columbia’s Data Catalogue. The existing road network inputted into the model was created by filtering out the forest tenure roads with an award date greater than 1990. Because the objective of the roads package is to project forest resource roads, the DRA roads were left untouched for the existing 1990 road network. The projection results would be compared with the observed modern road network which was simply a complete and unfiltered combination of the DRA and Forest Tenure road datasets, representing the most complete and up-to-date road network for the area.   
Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) Government of British Columbia’s Data Catalogue (Government of British Columbia, 2021b). Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested ‘reserved’ areas, and satellite image change detection may not always represent the entire cutblock area (Government of British Columbia, 2021b). Forest harvest cut blocks with harvest years greater than 1990 were filtered and used as targets for the projections.

@@

Projections were run in R (R Core Team, 2020) within the Revelstoke TSA. The forest road network projections were run using two different resolutions of the cost surface in order to assess how sensitive the results were to pixel size; the finer resolution used cost surface pixels of 100 meters by 100 meters and was the focus of this study, whereas the coarse resolution used an aggregated cost surface with pixels of 1000 meters by 1000 meters. The least cost path method yields projections that are obviously less realistic than the minimum spanning tree method (Figure 2), so we focussed on assessing the performance of the minimum spanning tree method and investigated sensitivity to sampling types (regular and random) and sampling densities (0.000001 for low and 0.0001 for high) (Figure 2).

In order to assess the performance of the projections, the following metrics were used to compare the projected forest road network results with the observed network. Road presence is a simple measure of whether or not roads are present in each pixel using the cost surface as a template. Road disturbance footprint was created as done by Ibisch et al. (2016) where a 1 km buffer was applied to all roads. Forestry disturbance footprint was created by buffering 500 metres around all roads and cutblocks. This metric relates to the anthropogenic disturbance metric used for boreal caribou (Johnson et al., 2020), and can be calculated using the caribouMetrics R package (Endicott et al.,in prep). The 100m x 100m resolution cost surface was used as the template to derive the forestry disturbance footprint metric values. Road density is an important metric that is often used when studying anthropogenic impacts on wildlife (Beazley et al., 2004). Road density is represented in meters per hectare (m/ha) and was calculated using the rasterizeLineDensity function within the caribouMetrics package (Endicott et al., in prep). This was done by intersecting the roads with the cost surface raster grid, calculating the length of road in each grid cell and dividing by the grid cell resolution. Distance to nearest road assigns each cell within the TSA a distance (meters) to the nearest road segment. The metric can be calculated using the “dist_to_nearest” function in the “starsExtra” package (Dorman, 2021), but this implementation can be prohibitively slow. We have implemented a faster moving window method in the XX function of the roads package.

Using the metrics described above, we assessed the aspatial accuracy of the projections by calculating the mean values of the projection metrics rasters and then compared them with the means of the observed metric rasters. This was done at an overall TSA level (where means were taken from across the entire study area) and at a cutblock level (where metric rasters were masked by the cutblock polygons and the means were taken only from within these areas).

The projections were assessed for their spatial accuracy in two different ways. First, difference maps were created by subtracting the road presence rasters of the projected networks with the observed network in order to further highlight discrepancies between the networks. The observed and projected road network shapefiles were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in GIS (QGIS) to gain a better understanding of what was happening on the ground. Secondly, a roaded/roadless matching analysis using the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was performed. This matching analysis compared the three metric rasters for all projection methods with the observed metric rasters and calculated percentages of the TSA which matched and didn’t match. All pixels in the Revelstoke TSA for the projected networks fell into 5 categories when compared with the observed raster: Roaded prior to 1990 is the roaded area of the existing 1990 road network that was inputted into the models to project future forestry roads; agree roadless is the roadless area where pixels between projected and observed match; agree roaded after 1990 is the roaded areas where projected forestry roads match with observed forestry roads; project roaded & observe roadless is areas where the projection was roaded but where the observed network was roadless; and finally project roadless & observe roaded represents areas where the projection was roadless but the observed network was roaded. Using these 5 categories, the landscape within the Revelstoke TSA was categorized for each projection method and summarized (figure 3).

# Results
## Aspatial accuracy
Various projection methods provided more accurate aspatial results than others when compared with the observed metrics depending on the metric used. The forestry disturbance footprint metric produced means across all projections that were closest with observed values (Figure 3). Projections of road disturbance footprint and distance to road were also reasonably accurate across the whole landscape, with distance to road being less accurate within cutblocks. Road presence and road density were less accurate and much more sensitive to assumptions about the distribution of roads within cutblocks (i.e. sample density). The road presence metric is sensitive to small differences in road location while the road density metric is sensitive to variations in the curves of tightly packed roads within grid cells that cannot be modeled using this algorithm. Road presence and road density were also the most sensitive to pixel size (reference appendix table for coarse resolution projections). The projection using the Clement’s (2021) QGIS plug-in also yielded quite accurate results, especially at an overall TSA level.

## Spatial accuracy
  Difference maps helped to highlight areas of discrepancy between the projections and the observed forestry road networks. In areas where the quality of the cutblock and road data was high, the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (figure 4a). Areas of greatest discrepancy often suggest problems with available forestry road or cutblock data. There are some areas in Revelstoke, for example, where there are mapped cutblocks but no mapped access roads, even though satellite images clearly show roads (Figure 4b). There are also instances where areas were likely misidentified as cutblocks (Figure 4c). This is likely due to difficulty distinguishing cutblocks from other unforested areas such as scree slopes and meadows using remote sensing methods.

@@

```{r get-data}
##### Metric table bar plots
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(ggpubr)
library(MetBrewer)
library(here)

data_path_drvd <- "analysis/data/derived_data/"

fig_path <- "analysis/figures/"

fig_dev <- "jpeg"

# Figure 3 #====================================================================

# data (from "meanTable") in the runProjections script
meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))

@@

```{r fig3, fig.height=7, fig.width=fig_widths["two"], fig.cap="Revelstoke TSA mean road metrics using different projection methods. Observed road network metric values are represented by the horizontal dotted lines for each metric. The “overall” column represents mean metric values from across the entire area of the TSA,  whereas the right “cutover” column represents mean values from within cutblock areas only."}
# organize data for ggplot
meanTable_long <- meanTable %>%
  pivot_longer(-c(sampleType, sampleDens, areaMean),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleDens = factor(sampleDens,
                             levels = c("centroid",
                                        "low sample density",
                                        "high sample density",
                                        "klementQGIS"),
                             labels = c("Centroid", "Low\ndensity", "High\ndensity",
                                        "Klement\nQGIS")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road density", "Road presence",
                                    "Distance to road",
                                    "Forestry disturbance footprint",
                                    "Road disturbance footprint")))

@@

Figure \@ref(fig:fig3) shows how we can have a caption and cross-reference for a plot

# Discussion
Through assessing the performance of our model by comparing its spatial and aspatial
accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters. However, for applications where the exact location of roads is important (like operational road engineering purposes), or for projecting impacts on wildlife that respond to roads at a very fine spatial scale then this is not an appropriate model. The model does, however, have potential to be a useful tool for predicting impacts on wildlife across large landscape simulations and could easily be incorporated as an element into more complex models that take into account other landscape phenomena such as wildfires, climate change, etc. Because this model does not take timescales into account, it could be a useful post-processing tool as an element of more complex models as it can be applied retroactively to landscape simulations.

>>>>

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300,
  cache = TRUE
)
```

@@

# Introduction

@@

Resource extraction, such as forest harvesting, is the main driver of road construction in much of Canada's forested area. In British Columbia, over 600 000 kilometers of resource roads make up the vast majority of the province's total road network (83%) (Environmental Reporting BC, 2018; Forest Practices Board, 2015). Over 75% of resource roads in the province were built by the forest industry (Forest Practices Board, 2015). Resource extraction activities provide jobs and renewable raw materials (Canada, 2013b), but they also can create threats for wildlife. For example, the main threats to woodland caribou, a species at risk in Canada, are human disturbances such as forest harvesting, oil and gas exploration/extraction, and road networks. These particular disturbances fragment their habitat, and create open areas and extensive young forests that attract other species such as moose and deer which in turn, attracts an increased number of predators, such as wolves and even poachers (Canada, 2013a). Woodland caribou are often considered to be an umbrella species, which means that protecting and restoring their habitat would benefit other species also dependent on old forests, such as fishers, pine martens, and many species of birds (Mulvihill, 2018).

@@

Forest simulation models generally either fall into two categories: deterministic and stochastic (Boychuk et al., 2009; Helmes & Stockbridge, 2011). Stochastic simulation approaches use probability distributions in combination with random number generators to determine events (He & Mladenoff, 1999). Despite the importance of roads on the environment, previous spatial stochastic harvest simulation models rarely include forest road network development in their projections. The "roads" package in R aims to fill this gap in landscape modelling by projecting large-scale forest road networks based on future harvest cut blocks, an existing road network, and a road cost surface. This paper uses the "roads" package to run forestry road network simulations in British Columbia from approximately 1990 to present and compares the results to the observed modern-day forest road networks in order to assess the performance of the package and to showcase its usefulness within forest simulations.

(Introduction round 2: road context) Several papers have previously introduced forest road network projections (Anderson & Nelson, 2004; Clark et al., 2000; Dean, 1997) however many of these methods are complex with many input parameters and are not readily available for public use. These previous projections have mainly been developed in order to facilitate operational forest road network planning and therefore require precise data and many input parameters in order to obtain the most effective network. These projection methods are therefore not very applicable for projecting forest roads over large-scale landscapes. On the other hand, many spatial stochastic forest management simulations haven't included forest road development at all (citations). One unpublished QGIS plugin by Clement (2021) is available for projecting forest roads and is also included in the analysis of this paper (GitHub - Klemet/ForestRoadNetworkPluginForQGIS: A plugin for QGIS that is able to construct a forest road network based on a cost matrix, polygons to reach, a current road network, the dijkstra algorithm and 3 network-creation heuristics.). There is therefore a need in large scale landscape simulations for a tool that can adequately project forest road development with less complexity and minimal input parameters that can easily be incorporated into other complex simulations in large scale landscapes. As many ecological forecasts use increasingly complex methods and data, forecasting tools (e.g., models) should also be scalable and leverage existing models (Bodner et al., in press) ("Bridging the divide between ecological forecasts and environmental decision-making").

@@

# Methods

## Study Area

We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as our study area because forestry is the dominant resource extraction activity in this region and much of the TSA's road network is composed of forestry resource roads built to gain access to forest harvest cutblocks (Figure 1). This allows us to project road building based on forest harvest data. With approximately 55% of Revelstoke TSA's area being non-forested land (alpine, lakes, swamp, brush, rock, etc) (Brown & Hachey, 2010), the TSA also provides an ideal location to observe the behaviour of the projections across a landscape which presents multiple barriers to road development. The Revelstoke TSA is in the interior-wet belt of the province and its forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) (Ministry of Forests, n.d.-b). The TSA contains wildlife species that are at risk due to their declining populations and are thought to be negatively affected by road development, such as members of the southern mountain woodland caribou population (Brown & Hachey, 2010). The Revelstoke TSA is an example of a landscape where road projections could be used to manage critical wildlife habitats. The British Columbia TSA boundary shapefile layer was downloaded from the British Columbia Data Catalogue (Government of British Columbia, 2021a).

```{r fig1, fig.cap="Revelstoke timber supply area in British Colombia, Canada. The colour shows the cost of building roads across each pixel."}
##### Metric table bar plots
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(ggpubr)
library(MetBrewer)
library(here)
library(tmap)
library(sf)

#set paths
data_path_raw <- "analysis/data/raw_data/"
data_path_drvd <- "analysis/data/derived_data/"

tsaBoundary <- read_sf(here(data_path_raw, "new_tsa27_boundaries.shp"))
#cost surface raster layer
bc_cost_surface <- terra::rast(here(data_path_raw, "cost_surface_bc_ha.tif"))
lakeValue <- 65000

tsaCost <- terra::crop(bc_cost_surface, tsaBoundary)
tsaCost <- terra::subst(tsaCost, from = NA, to = lakeValue)
tsaCost <- terra::mask(tsaCost, terra::vect(tsaBoundary))

canada <- geodata::gadm("CAN", level = 1, path = here(data_path_raw))

canada <- canada %>% 
  st_as_sf() %>% 
  filter(NAME_0 == "Canada") %>% 
  st_transform(st_crs(tsaBoundary)) %>% 
  st_simplify(dTolerance = 10000)

canada_overlay <- tm_shape(canada, projection = "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")+
  tm_borders(col="black")+
  tm_shape(st_buffer(tsaBoundary, 10000))+
  tm_polygons(border.col = met.brewer(pal_nm, 1), col = met.brewer(pal_nm, 1))+
  tm_layout(frame=TRUE)

tsa <- tm_shape(tsaCost, name = "Cost")+
  tm_raster(style = "cont", title = "Cost")+
  tm_shape(st_union(tsaBoundary))+
  tm_borders(col = "black")+
  tm_compass(position = c(0.3, 0.11), size = 1)+
  tm_scale_bar(position = c("center", "bottom"), text.size = 1)+
  tm_layout(legend.position = c("left", "bottom"),
            legend.bg.color = "white")

vp <- grid::viewport(x=0.965, y=0.945, width = 0.35, height=0.2,
               just=c("right", "top"))

tmap_save(tsa, filename = here("analysis/figures/Figure1_StudyArea.png"),
  dpi = 600, insets_tm = canada_overlay, insets_vp = vp,
  height = 6, width = fig_widths["single"], units = "in")

knitr::include_graphics(here("analysis/figures/Figure1_StudyArea.png"))

## Data

To spatially define the cost of constructing and traveling on a forestry road we developed a cost surface raster. We used several spatial raster layers to construct our cost surface. For all cells the base cost followed the Interior Appraisal Manual of BC (source) where the cost per kilometer of road was a linear function of slope. Separate costs for river and pipeline crossings were then added to the cost. Major water bodies were considered barriers and assigned a very high cost. Lastly, to coerce the algorithm to follow existing roads, all nodes representing the existing road network were assigned a cost of 0 (Table 1). The cost surface for the Revelstoke TSA consisted of 1703936 pixels with a resolution of 100 meters by 100 meters.

Table 1. Definitions of road building costs for building the cost surface.

```{r}
tbl <- "Cost Type|Road Attribute|Value ($ per km)|Source
Barriers of road development|Major lakes|65000|Major water bodies of BC
Impedance|Existing roads|0|Assumed
Road Construction Cost Structure|Pipe crossing|1911|Interior Appraisal Manual
|River crossing|2130|Interior Appraisal Manual
|Base cost|16178 + 504*slope|Interior Appraisal Manual
"
df <- read.delim(textConnection(tbl),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df) <- unname(as.list(df[1,]))
df <- df[-1,] # remove first row
row.names(df)<-NULL
knitr::kable(df)
```

Our road network data was obtained by combining the digital road atlas (DRA) (Data BC, 2021) and forest tenure road (FLNRORD, 2021b) data sets for BC. The existing road network, used as the starting point for projections, was created by filtering out the forest tenure roads with an award date greater than 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset (Government of British Columbia, 2021b). Cutblock information is compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area (Government of British Columbia, 2021b). Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

## Road Projection Methods

We developed the `roads` R package to project where new roads will be built based on the existing road network, locations of expected resource extraction (i.e. harvest cutblocks) and the landscape connecting the two. There are several different methods available for solving this problem, including snapping, least cost paths, and minimum spanning trees. Snapping follows a straight line between the existing road and each cutblock regardless of the intervening landscape. The least cost paths (LCP) method uses Dijkstra's shortest path algorithm (1959) to determine the optimal path between a cutblock and existing road based on a the cost of building a road at any point between the two. The cost is represented by a cost raster described below. Finally, the minimum spanning tree (MST) method builds on the LCP method by determining the least cost path between all cutblocks and the road and between each cutblock. Then the minimum spanning tree connecting all cutblocks to the road is solved using Kruskal's algorithm (1956). (See Appendix X for a more detailed explanation of these methods). In the remainder of this paper, we focus on the LCP and MST approaches given their assumptions are better aligned with how road networks are designed.

In addition to the projection methods included in the `roads` package we also applied the Forest Road Network Plugin for QGIS by Clément Hardy (<https://github.com/Klemet/ForestRoadNetworkPluginForQGIS>). This plugin uses a similar least cost path algorithm to the `roads` package but instead of building roads along the least cost path to each cutblock independently the Forest Road Network Plugin uses a heuristic to determine which cutblocks to connect first and then builds the roads sequentially. The default heuristic is to connect the closest cutblocks first. We supplied our cost surface, cutblocks and road network data and kept all the default settings within the QGIS plugin.

@@

We projected road development in the Revelstoke TSA from 1990 to the present using several different inputs and compared the projections to the observed present day road network in order to evaluate the accuracy of the projections and their sensitivity to different inputs. We compared two resolutions for the cost surface by aggregating the fine resolution cost raster to a coarser 1000 m resolution. We also compared methods for locating landing points with in cutblocks by comparing using the polygon centroid, regularly spaced and random points, as well as point densities of 1 and 100 pts/km^2^ (Figure \@ref(fig:fig2)).

```{r fig2, fig.height=5, fig.width=fig_widths["two"], fig.cap="Example of the different projection methods and sampling strategies."}
knitr::include_graphics(here("analysis/figures/projection methods figure.jpg"))
```

In order to assess the performance of the projections, the following metrics were used to compare the projected forest road network results with the observed network: 1) Road presence, a simple measure of whether or not roads are present in each pixel; 2) Road disturbance footprint, 1 km buffer applied to observed and projected roads (Ibisch et al., 2016); 3) Forestry disturbance footprint, a 500 m buffer applied to observed and projected roads and cutblocks; 4) Road density, density of roads in meters per hectare (m/ha); 5) Distance to nearest road, distance from each raster cell to the nearest road. The forestry disturbance footprint follows the anthropogenic disturbance metric used for boreal caribou (Johnson et al., 2020), and was calculated using the `caribouMetrics` R package (Endicott et al.,in prep). Road density is an important metric that is often used when studying anthropogenic impacts on wildlife (Beazley et al., 2004). Road density was calculated using the rasterizeLineDensity function within the `caribouMetrics` package (Endicott et al., in prep). Distance to nearest road was calculated using a fast moving window method in the `getDistFromSource` function of the roads package. Each metric was applied to the results from each projection method and the observed roads with the results were stored as rasters.

The aspatial accuracy of the projections was assessed by calculating the mean values of the projection metric rasters and then comparing them with the means of the observed metric rasters. This was done across the TSA and at the cutblock level, where metric rasters were masked by the cutblock polygons and the means were taken only from within cutblocks.

The projections were assessed for their spatial accuracy in two different ways. First, difference maps were created by subtracting the road presence rasters of the projected networks with the observed network in order to highlight discrepancies between the networks. The observed and projected road network shapefiles were then overlain on the difference maps and regions of major discrepancies were investigated using Google satellite imagery in QGIS to investigate potential causes of the discrepancy. Secondly, a matching analysis using the three binary metrics (road presence, road disturbance footprint, and forestry disturbance footprint) was performed. This matching analysis compared the three metric rasters for all projection methods with the observed metric rasters and calculated percentages of the TSA that matched and didn't match. All pixels in the Revelstoke TSA for the projected networks fell into 5 categories when compared with the observed raster: 1) pre-existing roads; 2) agree roadless, both projected and observed show no roads; 3) agree roaded, both projected and observed show new roads; 4) false positive, the projection showed a road but no roads were observed; and 5) false negative, the projection showed no roads but a road was observed. Using these 5 categories, the landscape within the Revelstoke TSA was categorized for each projection method and summarized.

# Results

## Aspatial accuracy

Various projection methods provided more accurate aspatial results than others when compared with the observed metrics depending on the metric used. The forestry disturbance footprint metric produced means across all projections that were closest with observed values (Figure 3). Projections of road disturbance footprint and distance to road were also reasonably accurate across the whole landscape, with distance to road being less accurate within cutblocks. Road presence and road density were less accurate and much more sensitive to assumptions about the distribution of roads within cutblocks (i.e. sample density). The road presence metric is sensitive to small differences in road location while the road density metric is sensitive to variations in the curves of tightly packed roads within grid cells that cannot be modeled using this algorithm. Road presence and road density were also the most sensitive to pixel size (reference appendix table for coarse resolution projections). The projection using the Clement's (2021) QGIS plug-in also yielded quite accurate results, especially at an overall TSA level.

## Spatial accuracy

Difference maps helped to highlight areas of discrepancy between the projections and the observed forestry road networks. In areas where the quality of the cutblock and road data was high, the projections remained within a reasonable vicinity of the observed roads and also realistically remained within valley bottoms without cutting across natural barriers such as mountains (figure 4a). Areas of greatest discrepancy often suggest problems with available forestry road or cutblock data. There are some areas in Revelstoke, for example, where there are mapped cutblocks but no mapped access roads, even though satellite images clearly show roads (Figure 4b). There are also instances where areas were likely misidentified as cutblocks (Figure 4c). This is likely due to difficulty distinguishing cutblocks from other unforested areas such as scree slopes and meadows using remote sensing methods.

@@

```{r get-data}
# load csv
meanTable <- read.csv(here(data_path_drvd, "mean_table.csv"))

matchData <- read.csv(here(data_path_drvd, "agree_table.csv"))

@@

```{r fig3, fig.height=7, fig.width=fig_widths["two"], fig.cap="Revelstoke TSA mean road metrics using different projection methods. Observed road network metric values are represented by the horizontal dotted lines for each metric. The “overall” column represents mean metric values from across the entire area of the TSA,  whereas the right “cutover” column represents mean values from within cutblock areas only."}
# organize data for ggplot
meanTable_long <- meanTable %>%
  pivot_longer(-c(sampleType, sampleDens, areaMean),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleDens = factor(sampleDens,
                             levels = c("centroid",
                                        "low sample density",
                                        "high sample density",
                                        "klementQGIS"),
                             labels = c("Centroid", "Low\ndensity", "High\ndensity",
                                        "Hardy\nQGIS")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road density", "Road presence",
                                    "Distance to road",
                                    "Forestry disturbance footprint",
                                    "Road disturbance footprint")))

@@

```{r fig4, fig.height=5, fig.width=fig_widths["two"], fig.cap="Spatial accuracy"}

stackedBar <- matchData %>% ungroup() %>%
  mutate(agreement = factor(agreement,
                            levels = c("False negative", "False positive",
                                       "Agree roaded","Agree roadless",
                                       "Pre-existing roads")),
         sampleDens = paste(sampleType, sampleDens) %>%
           str_replace("1e-06", "low density") %>%
           str_replace("1e-04", "high density") %>%
           str_replace("centroid high density", "centroid") %>%
           str_replace("klementQGIS NA", "klementQGIS") %>%
           factor(levels = c("centroid",
                             "random low density",
                             "random high density",
                             "regular low density",
                             "regular high density",
                             "klementQGIS"),
                  labels = c("Centroid", "Random low density", "Random high density",
                             "Regular low density",
                             "Regular high density",
                             "Hardy QGIS")),
         metric = factor(metric,
                         levels = c("roadPresence",
                                    "forestryDisturbance",
                                    "roadDisturbance"),
                         labels = c("Road presence",
                                    "Forestry disturbance footprint",
                                    "Road disturbance footprint"))) %>%
  group_by(metric, sampleDens, agreement) %>%
  summarise(count = sum(count), .groups = "drop_last") %>%
  mutate(perc = count/sum(count)*100) %>% 
  ggplot(aes(x = sampleDens, y =perc, fill = agreement))+
  geom_col(position = "stack", width = 0.75)+
  theme_classic()+
  theme(axis.text.x=element_text(color = "black", size=11, angle=50, vjust=1, hjust=1))+
  scale_fill_manual(values = c(met.brewer(pal_nm, 4, direction = -1)[c(1,4,3,2)],
                               "#999999"))+
  theme(text = element_text(size = 15), axis.title.x = element_blank())+
  theme(legend.position="bottom", legend.text = element_text(size = 10), 
        legend.title = element_blank(),
        strip.background = element_blank())+
  labs(y = "Percent of Landscape")+
  facet_wrap(~metric)

stackedBar

# Discussion
